---
title: "Untitled"
author: "Anna Perez Welter"
date: "1/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
library(here)
```

```{r}
ca_dem <- read_csv(here("data", "ca_census_demographics_2010.csv"))
ca_pb <- read_csv(here("data", "ca_pollution_burden.csv"))
```

# Data wrangling
```{r}
ca_pb_nopct <- ca_pb %>%
  clean_names() %>%
  select(-contains("perc")) %>%
  select(-contains("pctl")) %>%
  select(-latitude, -longitude)
#removes unwanted categories
```

```{r}
ca_dem_clean <-ca_dem %>%
  clean_names()

```

## 3. Pollution Burden Indicators Only

```{r}
ca_pb_subset <- ca_pb_nopct %>%
  select(ozone: solid_waste, asthma:housing_burden)
```

Now for the PCA
Us the 'prcomp()' function
```{r, eval= FALSE}
pb_pca <- prcomp(ca_pb_subset, scale = TRUE) #error means that there is missing data
```

Aside: Exploring missingness. 
```{r}
summary(ca_pb_subset)
#NAs are relatively low, so it could be helpful not harmful to exclude. 
gg_miss_var(ca_pb_subset) #gives a bar graph of each category's missing data
matrixplot(ca_pb_subset, sortby = "poverty")
#missingness is shown as a red spot, darkness indicates data density
```
Remove (by complete case deletion) any observation with ANY 'NA' values.
```{r}
ca_pb_nona <- ca_pb_subset %>%
  drop_na()
summary(ca_pb_nona) #confirm that all NA values are removed. 

skim(ca_pb_nona) #nicely formatted summary table
```

Now there are no 'NA' values, so we can do PCA. (Otherwise consider an imputation package.)

```{r}
my_ca_pca <- prcomp(ca_pb_nona, scale= TRUE)

my_ca_pca
summary(my_ca_pca)
#show the proportion of variants explained by each component

biplot(my_ca_pca) #variables in red

# Alternative, cleaner plot:
my_biplot <- autoplot(my_ca_pca, 
                      colour = NA, # not showing the location of points
                      loadings.label.size = 3, 
                      loadings.label.colour ="black", 
                      loadings.label.repel = TRUE) 
my_biplot
```

##4. PCA for pollution burden + demographics. 

Join together ca_dem_clean with ca_pb_nopct via census tract category (has different names). 
```{r}
#inner join: keep the parts of the dataset that have a complete match

ca_df <- ca_dem_clean %>%
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract")) #tells it which two things to match

#drop NAs by complete case deletion: 
ca_nona <- ca_df %>%
  drop_na()
                              
```

Make a subset of ca_nona that only includes: 
-white-percent
-elderly_65_percent
-pm2_5
-pesticides
-traffic
-asthma
-cardiovascular disease
-poverty

```{r}
my_sub <- ca_nona %>%
  select(white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

PCA
```{r}
my_dem_pca <-prcomp(my_sub, scale= TRUE)

my_dem_biplot <- autoplot(my_dem_pca, 
                      colour = NA, # not showing the location of points
                      loadings.label.size = 3, 
                      loadings.label.colour ="black", 
                      loadings.label.repel = TRUE) 
my_dem_biplot
```

